---
title: "Homework 03 - Data Analysis with Matrix Algebra"
author: "Alessandro Corradini - Harvard Data Science for Life Science XSeries"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Matrix Algebra Examples Exercises

Suppose we are analyzing a set of 4 samples. The first two samples are from a treatment group A and the second two samples are from a treatment group B. This design can be represented with a model matrix like so:

```{r}
X <- matrix(c(1,1,1,1,0,0,1,1),nrow=4)
rownames(X) <- c("a","a","b","b")
```

X will then look like:

```
a 1 0
a 1 0
b 1 1
b 1 1
```

Suppose that the fitted parameters for a linear model give us:

```{r}
beta <- c(5, 2)
```

Use the matrix multiplication operator, %*%, in R to answer the following questions:

### Matrix Algebra Examples Exercises #1

What is the fitted value for the A samples? (The fitted Y values.)

```{r}
fitted = X %*% beta

fitted[ 1:2, ]
```

### Matrix Algebra Examples Exercises #2

What is the fitted value for the B samples? (The fitted Y values.)

```{r}
fitted = X %*% beta

fitted[ 3:4, ]
```

Suppose now we are comparing two treatments B and C to a control group A, each with two samples. This design can be represented with a model matrix like so:

```{r}
X <- matrix(c(1,1,1,1,1,1,0,0,1,1,0,0,0,0,0,0,1,1),nrow=6)
rownames(X) <- c("a","a","b","b","c","c")
```

which results in a matrix that looks like

```
a 1 0 0
a 1 0 0
b 1 1 0
b 1 1 0
c 1 0 1
c 1 0 1
```

Suppose that the fitted values for the linear model are given by:

```{r}
beta <- c(10,3,-3)
```

### Matrix Algebra Examples Exercises #3

What is the fitted value for the B samples?

```{r}
fitted = X %*% beta

fitted[ 3:4, ]
```

#### Matrix Algebra Examples Exercises #4

What is the fitted value for the C samples?

```{r}
fitted = X %*% beta

fitted[ 5:6, ]
```