---
title: "Homework 02 - Introduction to Matrix Albera"
author: "Alessandro Corradini - Harvard Data Science for Life Science XSeries"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Matrix Notation Exercises

### Matrix Notation Exercises #1

In R we have vectors and matrices. You can create your own vectors with the function c.

```{r}
c(1,5,3,4)
```

They are also the output of many functions such as

```{r}
rnorm(10)
```

You can turn vectors into matrices using functions such as rbind, cbind or matrix.

Create the matrix from the vector 1:1000 like this:

```{r}
X = matrix(1:1000,100,10)
```

What is the entry in row 25, column 3 ?

```{r}
X[25,3]
```


### Matrix Notation Exercises #2

Using the function cbind, create a 10 x 5 matrix with first column x=1:10. Then columns 2*x, 3*x, 4*x and 5*x in columns 2 through 5.

What is the sum of the elements of the 7th row?

```{r}
x = 1:10

X=cbind(x, 2*x, 3*x, 4*x, 5*x)

sum(X[7,])
```
### Matrix Notation Exercises #3

Which of the following creates a matrix with multiples of 3 in the third column?

- matrix(1:60,20,3)
- **matrix(1:60,20,3,byrow=TRUE)**
- x=11:20;rbind(x,2*x,3*x)
- x=1:40;matrix(3*x,20,2)


## Matrix Operation Exercises

### Matrix Operation Exercises #1
Suppose X is a matrix in R. Which of the following is NOT equivalent to X?


- t( t(X) )
- **X %*% matrix(1,ncol(X) )**
- X*1
- X%*%diag(ncol(X))

### Matrix Operation Exercises #2

Solve the following system of equations using R:

3a + 4b - 5c + d = 10

2a + 2b + 2c - d = 5

a -b + 5c - 5d = 7

5a + d = 4

What is the solution for c:

```{R}
X = matrix(c(3,2,1,5,4,2,-1,0,-5,2,5,0,1,-1,-5,1),4,4)

y = c(10,5,7,4)

sol = solve(X,y)

sol[ 3 ]
```

Load the following two matrices into R:

```{r}
a <- matrix(1:12, nrow=4)
b <- matrix(1:15, nrow=3)
```

Note the dimension of 'a' and the dimension of 'b'.

In the question below, we will use the matrix multiplication operator in R, %*%, to multiply these two matrices.

### Matrix Operation Exercises #3

What is the value in the 3rd row and the 2nd column of the matrix product of 'a' and 'b'

```{r}
d = a %*% b

d[ 3,2 ]
```

### Matrix Operation Exercises #4

Multiply the 3rd row of 'a' with the 2nd column of 'b', using the element-wise vector multiplication with *.

What is the sum of the elements in the resulting vector?

```{r}
sum(a[ 3, ] * b[ ,2 ])
```